### 2014-05-09 ApTranscriptome_TR.Rmd
Completed analysis to check temp at which expression up-regulated for High genes

1. describe the *de novo* assembly of the transcriptome for two ant colonies with in the *Aphaenogaster rudis-picea-fulva* species complex `r citep("10.1155/2012/752815")`
2. identify thermally-responsive genes
3. perform gene set enrichment analysis

@@

Two ant colonies were used for the transcriptome sequencing. The first, designated A22, was collected at Molly Bog, Vermont in August 2012 by Nick Gotelli and Andrew Nguyen. This colony was putatively identifed as *A. picea*. The second colony, designated Ar, was collected by Lauren Nichols in Raleigh, North Carolina. These colonies were maintained in the lab for 6 months prior to sample collection.

@@

# calculate overall P value for 'model' for each transcript
RxNpval <- ddply(TPM.dt.sub, .(Transcript), .inform="TRUE", modpFunc)
```

@@

```{r plotthermalbreadth, echo=FALSE, eval=TRUE}
# prep data for plotting
Ap.int.sd <- rbind(A22.int.sd, Ar.int.sd)

# ggplot
g1 <- ggplot(Ap.int.sd, aes(x=exp_sd, fill=colony)) + geom_density(alpha=0.2, position="identity")
g1 + scale_y_continuous(name="Density") +
  scale_x_continuous(name=expression("Standard deviation of expression"))
```

@@

png(paste(resultsdir, "PDF_expression_all.png", sep=""))
g3 <- ggplot(Ap.df, aes(x=max.val, fill=colony)) + 
  geom_density(alpha=0.2, position="identity") + 
  #scale_fill_manual(name = "Colony", values = c("white", "black")) + 
  scale_y_continuous(name="Density") +
  scale_x_continuous(name="Temperature of maximum expression")
suppressWarnings(print(g3))
dev.off()
```

@@

# apply predFunc to responsive transcripts
testpred <- ddply(resp.TPM.dt.sub[1:220], .(Transcript), .inform="TRUE", predFunc)

# calculate temperature of greatest change
T.on.df <- ddply(testpred, .(Transcript, colony), summarise, 
              T_on <- diff(pTPM))

testdiff[(which(diff(testdiff$pTPM) == max(diff(testdiff$pTPM)))+1), ]

```

@@

```{r echo=FALSE, results='hide'}
responsive.lms.ann.type <- signif.transcripts[names(responsive.lms)]
# combine responsive.lms.ann with "type" information
responsive.lms.ann.type <- merge(responsive.lms.ann.type, Ap.response.type[,c("Transcript", "A22.type", "Ar.type")], by = "Transcript", all = TRUE)
str(responsive.lms.ann.type)
write.csv(responsive.lms.ann.type, file = "results/Ap_responsive_genes_20140409.csv", row.names = FALSE)
```

>>>>

1. describe the *de novo* assembly of the transcriptome for two ant species within the *Aphaenogaster rudis-picea-fulva* species complex `r citep("10.1155/2012/752815")`
2. identify thermally-responsive genes
3. evaluate differences in the expression patterns between the two species
3. perform gene set enrichment analysis of thermally-responsive genes for the two species

@@

Two ant colonies were used for the transcriptome sequencing. The first, designated *A22*, was collected at Molly Bog, Vermont in August 2012 by Nick Gotelli and Andrew Nguyen. The second colony, designated *Ar*, was collected by Lauren Nichols in Raleigh, North Carolina. These colonies were maintained in the lab for 6 months prior to sample collection. Bernice Bacon DeMarco (Michigan State University) identified colony *A22* as *A. picea* and *Ar* as *A. carolinensis*.

@@

# calculate overall P value and R^2 for each transcript
RxNpval <- ddply(TPM.dt.sub, .(Transcript), .inform="TRUE", modpFunc)
```

@@

```{r plot_thermal_breadth, echo=FALSE, eval=TRUE}
# prep data for plotting
Ap.int.sd <- rbind(A22.int.sd, Ar.int.sd)

# plot thermal breadth
g1 <- ggplot(Ap.int.sd, aes(x=exp_sd, fill=colony)) + geom_density(alpha=0.2, position="identity")
g1 + scale_y_continuous(name="Density") +
  scale_x_continuous(name=expression("Standard deviation of expression"))
```

@@

maxexplot <- ggplot(Ap.df, aes(x=max.val, fill=colony)) + 
  geom_density(alpha=0.2, position="identity") + 
  #scale_fill_manual(name = "Colony", values = c("white", "black")) + 
  scale_y_continuous(name="Density") +
  scale_x_continuous(name="Temperature of maximum expression")
suppressWarnings(print(maxexplot))
```

@@

# apply predFunc to all responsive transcripts
resp.TPM.dt.sub.pred <- ddply(resp.TPM.dt.sub, .(Transcript), .inform="TRUE", predFunc)

# setkey to Transcript and colony 
resp.TPM.dt.sub.pred <- data.table(resp.TPM.dt.sub.pred)
setkey(resp.TPM.dt.sub.pred, Transcript, colony)

# transcripts expressed at *High* temperatures in A22
A22.high.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(A22.high, "A22")]
str(A22.high.TPM.dt.sub)

# make data.frame for results
A22.high.T_on <- data.frame(Transcript = unique(A22.high.TPM.dt.sub$Transcript), colony = rep(A22.high.TPM.dt.sub$colony, length = length(unique(A22.high.TPM.dt.sub$Transcript))), T_on = NA, pT_on = NA)

# loop across transcripts, calculating T_on

for(i in unique(A22.high.TPM.dt.sub$Transcript)) {
    subdf <- A22.high.TPM.dt.sub[i]
    subdf <- subdf[which(subdf$val > 14), ]
    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
    pT_on <- subdf[median(which(diff(subdf$pTPM) == max(diff(subdf$pTPM))))+1, val]
    A22.high.T_on[which(A22.high.T_on$Transcript == i), "T_on"] <- T_on
    A22.high.T_on[which(A22.high.T_on$Transcript == i), "pT_on"] <- pT_on
}

# repeat for Ar
Ar.high.TPM.dt.sub <- resp.TPM.dt.sub.pred[J(Ar.high, "Ar")]
str(Ar.high.TPM.dt.sub)

Ar.high.T_on <- data.frame(Transcript = unique(Ar.high.TPM.dt.sub$Transcript), colony = rep(Ar.high.TPM.dt.sub$colony, length = length(unique(Ar.high.TPM.dt.sub$Transcript))), T_on = NA, pT_on = NA)

for(i in unique(Ar.high.TPM.dt.sub$Transcript)) {
    subdf <- Ar.high.TPM.dt.sub[i]
    subdf <- subdf[which(subdf$val > 14), ]
    T_on <- subdf[median(which(diff(subdf$TPM) == max(diff(subdf$TPM))))+1, val]
    pT_on <- subdf[median(which(diff(subdf$pTPM) == max(diff(subdf$pTPM))))+1, val]
    Ar.high.T_on[which(Ar.high.T_on$Transcript == i), "T_on"] <- T_on
    Ar.high.T_on[which(Ar.high.T_on$Transcript == i), "pT_on"] <- pT_on
}

cor.test(Ar.high.T_on$T_on, Ar.high.T_on$pT_on)
cor.test(A22.high.T_on$T_on, A22.high.T_on$pT_on)

t.test(Ar.high.T_on$T_on, A22.high.T_on$T_on)
t.test(Ar.high.T_on$pT_on, A22.high.T_on$pT_on)
```

Genes with increased expression at *High* temperatures are turned on at higher temperatures in *ApVT* than *AcNC*.

```{r plot_high_T_on, echo=FALSE, eval=TRUE}
# prep data for plotting
Ap.high.T_on <- rbind(A22.high.T_on, Ar.high.T_on)

# plot
png("T_on_high.png")
T_on_plot <- ggplot(Ap.high.T_on, aes(x=T_on, fill=colony)) + geom_density(alpha=0.2, position="identity")
T_on_plot + scale_y_continuous(name="Density") +
  scale_x_continuous(name=expression("Temperature"))
dev.off()

# plot predicted
png("pT_on_high.png")
pT_on_plot <- ggplot(Ap.high.T_on, aes(x=pT_on, fill=colony)) + geom_density(alpha=0.2, position="identity")
pT_on_plot + scale_y_continuous(name="Density") +
  scale_x_continuous(name=expression("Temperature"))
dev.off()
```

@@

```{r echo=FALSE, results='hide'}
responsive.lms.ann.type <- signif.transcripts[names(responsive.lms)]
# combine responsive.lms.ann with "type" information
responsive.lms.ann.type <- merge(responsive.lms.ann.type, Ap.response.type[,c("Transcript", "Ar.type", "Ar.type")], by = "Transcript", all = TRUE)
str(responsive.lms.ann.type)
write.csv(responsive.lms.ann.type, file = "results/Ap_responsive_genes_20140409.csv", row.names = FALSE)
```
